Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-01-06 paper/ms.tex
Modified notation in the appendix.

There are three primary applications we envision for the \genrp formalism: 1). modeling
the variability of an astrophysical system to infer it properties; 2).
accounting for astrophysical variability as a source of noise when trying to detect 
additional phenomena; 3). interpolating or extrapolat variability to future or missing
times.  Some examples of the first are determining the characteristic
variability timescale of quasars, measuring the asteroseismic variations of stars,
or detection of quasi-periodic variability in a high-energy source.  Examples of the
second are detecting transiting exoplanets and correcting for stellar activity in
radial velocity measurements. Examples of the third are measuring time delays in 
multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.

@@

Our background is in studying transiting exoplanets, a field which has only recently begun
to adopt full covariance matrices in analyzing the noise in transiting planet lightcurves.
One promising early attempt at accounting for correlated noise was the wavelet approach
of Carter \& Winn (2010).  Their technique allowed for a power-law spectrum for the
noise scaling as $\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate, description of stellar variability.  Their approach runs in
$\cal{O}(N)$ time for $\alpha=1$, which is `flicker' or pink noise;  however, this
particular type of noise overpredicts the power spectrum at low frequency compared
to the granulation power spectrum.  It also cannot describe quasi-periodic noise.

Further progress was made by parameterizing the covariance matrix with simple,
analytic functions that describe the autocorrelation function.  The parameters
of these functions can then be optimized to best match the observed variability
pattern of the residuals (after subtracting a transit model).  The disadvantage of
this approach is that the functions used are chosen somewhat arbitrarily (e.g.\
the exponential-squared function), and the Cholesky decomposition of the covariance
matrix for computing the likelihood takes $\cal{O}(N^3)$ operations which prohibits 
application of this technique to large datasets.

Given the drawbacks of these approaches, the \genrp formalism allows both a fast
computation of the likelihood in $\cal{O}(N)$ time, as well as a functional form
that accurately describes stellar variability due to the relation to the simple
harmonic oscillator power spectrum, which is an analog of stellar asteroseismic
oscillations.  As higher signal-to-noise observations of transiting exoplanet systems
are obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters, and so we expect that \genrp
will be important for transit timing, transit spectroscopy, Doppler beaming,
phase functions, and more.

@@

For the complex values with $d_j \ne 0$, the equations are modified as follows.  We introduce complex
auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
r_k^R - i r_k^I$, with $k \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric
set of equations, and since the values of these variables are discarded in the end, their sign
is irrelevant).  The length of the vectors $l_k$ and $r_k$
is the number of kernel components, $J$;  we denote the individual elements of these vectors
as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,J\}$.  Note that the first $J_0$ elements we take
to be the kernels with $d_j=0$, and for these we do not include the imaginary component
equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
components as well:  $\gamma_k = \left[\exp{(-(c_1+id_1) t_{k,k+1})} ... \exp{(-(c_J+id_J) t_{k,k+1})}\right]^T
 = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-c_j t_{k,k+1}}
\cos{(d_j t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-c_j t_{k,k+1}}\sin{(d_j t_{k,k+1})}$.
Equation (61) becomes:
\begin{equation} \label{Amb61}
% See 9/2/2016 notes
% Use n rather than k for {1...N} - use g, h & u, v?
% \gamma & \phi - real & imaginary 
\sum_{j=1}^J \left(\frac{a_j}{2} l_{k,j}^R +\frac{b_j}{2} l_{k,j}^I\right)+ \frac{d}{2} x_k + \sum_{j=1}^J \left[
\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{y_k}{2},
\end{equation}
for $j \in \{1,...,J\}$ and $k \in \{1,...,N\}$, where $h_j$ is the $j$th data point.  There is no imaginary component to equation (61)
in \citet{Ambikasaran:2015}.

Equation (60) in \citet{Ambikasaran:2015} has real and imaginary components:
\begin{eqnarray} \label{Amb60}
\gamma_{k,j}^R l_{k,j}^R +\gamma_{k,j}^I l_{k,j}^I + \gamma_{k,j}^R x_k - l_{k+1,j}^R &=& 0,\\
\gamma_{k,j}^I l_{k,j}^R -\gamma_{k,j}^R l_{k,j}^I + \gamma_{k,j}^I x_k + l_{k+1,j}^I &=& 0,
\end{eqnarray}
with $j \in \{1,...,J\}$ and $k \in \{1,...,N-1\}$
\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran:2015}.

Finally, equation (59) in \citet{Ambikasaran:2015} has real and imaginary components:
\begin{align} \label{Amb59}
-&r_{k,j}^R + \frac{a_j}{2} x_k + \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
 &r_{k,j}^I + \frac{b_j}{2} x_k + \gamma_{k,j}^I r_{k+1,j}^R - \gamma_{k,j}^R r_{k+1,j}^I &&=0,
\end{align}
with $j \in \{1,...,J\}$ and $k \in \{2,...,N\}$.  Note that for the $J_0$ real components
we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
imaginary components of equations (59) and (60) in \citet{Ambikasaran:2015} as well for these, which reduces the total number
of equations to solve.

@@

As with the GRP method, we can build an extended matrix which is real and symmetric, and which may be
solved with a banded solver, as in \citet{NumericalRecipes}.  The determinant of this extended matrix
may be computed from the LU decomposition, and it is related to the determinant of the kernel by a
factor of $2^N$ (this is due to dividing each equation by $2$ to yield a symmetric matrix).
We arrange the auxiliary variables with the $d_j=0$
terms first, followed by the complex variables, yielding the vector
$x_{ex} = [x_1, \{r_{2,j}^R,l_{2,j}^R;j=1,...,J_0\}, \{r_{2,j}^R,r_{2,j}^I,l_{2,j}^R,l_{2,j}^I;j=J_0+1,...,J\},
x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [h_1/2, \{0; j=1,...,J_0+2(J-J_0)\}, h_2/2, ..., h_N/2]$.
The extended matrix, $A_{ex}$, is constructed by inserting the coefficients of equations \ref{Amb61},
\ref{Amb60}, and \ref{Amb59} into a matrix of dimensions $(2J_0+4(J-J_0)+5) \times N_{ex}$.
The solution is found by solving $A_{ex} x_{ex} = y_{ex}$ with LU decomposition and back-substitution
with a banded algorithm.

>>>>

There are three primary applications we envision for the \genrp\ formalism: 1).\ modeling
the variability of an astrophysical system to infer it properties; 2).\
accounting for astrophysical variability as a source of noise when trying to detect 
additional phenomena; 3).\ interpolating or extrapolating variability to future or missing
times.  Some examples of the first are determining the characteristic
variability timescale of a quasar, measuring the asteroseismic variation of a star,
or detection of quasi-periodic variability in a high-energy source.  Examples of the
second are detecting transiting exoplanets and correcting for stellar activity in
radial velocity measurements. Examples of the third are measuring time delays in 
multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.

@@

Our background is in studying transiting exoplanets, a field which has only recently begun
to adopt full covariance matrices in analyzing the noise in transiting planet lightcurves.
One promising early attempt at accounting for correlated noise was the wavelet approach
of Carter \& Winn (2010).  Their technique allowed for a power-law spectrum for the
noise scaling as $\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate, description of stellar variability.  Their approach runs in
${\cal O}(N)$ time for $\alpha=1$, which is `flicker' or pink noise;  however, this
particular type of noise overpredicts the power spectrum at low frequency compared
to the granulation power spectrum.  It also cannot describe quasi-periodic noise.

Further progress was made by parameterizing the covariance matrix with simple,
analytic functions that describe the autocorrelation function.  The parameters
of these functions can then be optimized to best match the observed variability
pattern of the residuals (after subtracting a transit model).  The disadvantage of
this approach is that the functions used are chosen somewhat arbitrarily (e.g.\
the exponential-squared function), and the Cholesky decomposition of the covariance
matrix for computing the likelihood takes ${\cal O}(N^3)$ operations which prohibits 
application of this technique to large datasets.

Given the drawbacks of these approaches, the \genrp\ formalism allows both a fast
computation of the likelihood in ${\cal O}(N)$ time, as well as a functional form
that accurately describes stellar variability due to the relation to the simple
harmonic oscillator power spectrum, which is an analog of stellar asteroseismic
oscillations.  As higher signal-to-noise observations of transiting exoplanet systems
are obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters, and so we expect that \genrp\
will be important for transit timing, transit spectroscopy, Doppler beaming,
phase functions, and more.

@@

For the complex values with $d_j \ne 0$, the equations are modified as follows.  We introduce complex
auxiliary vectors $l_n = g_n - i h_n$, with $n \in \{1,2,...,N\}$ and $l_1 = 0$; $r_n =
u_n - i v_n$, with $n \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric
set of equations, and since the values of these variables are discarded in the end, their sign
is irrelevant).  The length of the vectors $l_n$ and $r_n$
is the number of kernel components, $J$;  we denote the individual elements of these vectors
as $g_{n,j}+ih_{n,j}$ and $u_{n,j}+iv_{n,j}$, $j \in \{1,...,J\}$.  Note that the first $J_0$ elements we take
to be the kernels with $d_j=0$, and for these we do not include the imaginary component
equations, which are unnecessary and only add extra computation time.  We also define vectors $\phi$ and $\psi$ with
complex components as well:  $\gamma_n = \phi_n + i\psi_n = \left[\exp{(-(c_1+id_1) t_{n,n+1})} ... \exp{(-(c_J+id_J) t_{n,n+1})}\right]^T
$.  The individual elements are given by $\phi_{n,j} = e^{-c_j t_{n,n+1}}
\cos{(d_j t_{n,n+1})}$ and $\psi_{n,j} = -e^{-c_j t_{n,n+1}}\sin{(d_j t_{n,n+1})}$.
Equation (61) becomes:
\begin{equation} \label{Amb61}
% See 9/2/2016 notes
% Use n rather than k for {1...N} - use g, h & u, v?
% \gamma & \phi - real & imaginary 
\sum_{j=1}^J \left(\frac{a_j}{2} g_{n,j} +\frac{b_j}{2} h_{n,j}\right)+ \frac{d}{2} x_n + \sum_{j=1}^J \left[
\phi_{n,j} u_{n+1,j} + \psi_{n,j} v_{n+1,j}\right] = \frac{y_n}{2},
\end{equation}
for $j \in \{1,...,J\}$ and $n \in \{1,...,N\}$, where $y_j$ is the $j$th data point.  There is no imaginary component to equation (61)
in \citet{Ambikasaran:2015}.

Equation (60) in \citet{Ambikasaran:2015} has real and imaginary components:
\begin{eqnarray} \label{Amb60}
\phi_{n,j} g_{n,j} +\psi_{n,j} h_{n,j} + \phi_{n,j} x_n - g_{n+1,j} &=& 0,\\
\psi_{n,j} g_{n,j} -\phi_{n,j} h_{n,j} + \psi_{n,j} x_n + h_{n+1,j} &=& 0,
\end{eqnarray}
with $j \in \{1,...,J\}$ and $n \in \{1,...,N-1\}$
\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran:2015}.

Finally, equation (59) in \citet{Ambikasaran:2015} has real and imaginary components:
\begin{eqnarray} \label{Amb59}
-u_{n,j} + \frac{a_j}{2} x_n + \phi_{n,j} u_{n+1,j} + \psi_{n,j} v_{n+1,j} &=&0,\\
 v_{n,j} + \frac{b_j}{2} x_n + \psi_{n,j} u_{n+1,j} - \phi_{n,j} v_{n+1,j} &=&0,
\end{eqnarray}
with $j \in \{1,...,J\}$ and $n \in \{2,...,N\}$.  Note that for the $J_0$ real components
we can drop the variables $v_{n,j}$ and $h_{n,j}$ as these are zero, and we can drop the
imaginary components of equations (59) and (60) in \citet{Ambikasaran:2015} as well for these, which reduces the total number
of equations to solve.

@@

As with the GRP method, we can build an extended matrix which is real and symmetric, and which may be
solved with a banded solver, as in \citet{NumericalRecipes}.  The determinant of this extended matrix
may be computed from the LU decomposition, and it is related to the determinant of the kernel by a
factor of $2^N$ (this is due to dividing each equation by $2$ to yield a symmetric matrix).
We arrange the auxiliary variables with the $d_j=0$
terms first, followed by the complex variables, yielding the vector
$x_{ex} = [x_1, \{u_{2,j},g_{2,j};j=1,...,J_0\}, \{u_{2,j},v_{2,j},g_{2,j},h_{2,j};j=J_0+1,...,J\},
x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [y_1/2, \{0; j=1,...,J_0+2(J-J_0)\}, y_2/2, ..., y_N/2]$.
The extended matrix, $A_{ex}$, is constructed by inserting the coefficients of equations \ref{Amb61},
\ref{Amb60}, and \ref{Amb59} into a matrix of dimensions $(2J_0+4(J-J_0)+5) \times N_{ex}$.
The solution is found by solving $A_{ex} x_{ex} = y_{ex}$ with LU decomposition and back-substitution
with a banded algorithm.
